# 11.6.1 SIMPLEST COMMUNITY OCCUPANCY MODEL: N-FOLD SINGLE SPECIES OCCUPANCY MODEL WITH SPECIES TREATED AS FIXED EFFECTS

This model is a straightforward site-occupancy model with one component model per observed species. Hence, species identity is retained, so the model accommodates features of the individual member species of a (meta)community as well as emerging characteristics of the metacommunity. Furthermore, heterogeneity among species in occurrence and detection can be fully accounted for. The simplest such model consists of fitting separate occupancy model parameters to each observed species, but inside a single model. No relationship among species will be imposed on the parameters. In ANOVA terms, species will be treated as fixed effects; the estimates for one species will not be affected by the data from another species. This analysis allows one to estimate (correcting for imperfect detection) the number of species occurring at each site among the total number of species ever detected in the metacommunity, but not the total number of species occurring in the entire metacommunity, because some species may have been missed everywhere. This is the approach taken in calculating the popular wildlife picture index (O'Brien et al., 2010).

```{r}
source("sandbox/11_6_setup.R")
```

This first and simplest, but already very powerful, approach is to simply fit a separate species distribution model (SDM) to every species. Of course, we will use our favorite SDM, the occupancy model from Chapter 10. We could do this in a loop by fitting a separate occupancy model to the data from each of the 145 species in turn. However, it is much more efficient to fit these 145 models to the data of all species at once. This can easily be done in BUGS by organizing the data as a three-dimensional array (which we have already done). For illustration, we simply fit a null model to each species, with a species-specific constant intercept for both the occupancy and the detection probability, though we could readily add covariates. One big advantage of this approach over one that fits 145 separate models is that in the combined model, we can easily sum or compare quantities across species. For instance, it is trivial to add up the estimated number of occurring species for each site with a full propagation of all uncertainties involved in producing such a sum. As usual, the ease with which we can do computations on latent variables (here, the indicators of occurrence z) is one of the great practical benefits of a Bayesian analysis using MCMC. Similarly, we could constrain parameters among species, for instance by making them constant for some or all species or modeling them as a function of species-specific covariates (as we will see in Section 11.6.3).

We start the series of multispecies or community occupancy models without accounting for any structure among the three replicate surveys, so that we can collapse the detection/nondetection data to detection frequencies (which tell us whether a species was detected 0, 1, 2, or 3 times). We fit the following two-level HM to our detection frequency data ysumik for species k (k ¼ 1.145) and site i (i ¼ 1.267).

```{r}

```
