The below code is adapted from: Applied Hierarchical Modeling in Ecology: Analysis of distribution, abundance and species richness in R and BUGS: Volume 1: Prelude and Static Models. Kéry and Royle 2016.

10.4 A SLIGHTLY MORE COMPLEX SITE-OCCUPANCY MODEL WITH COVARIATES

We will hardly ever use the null/null site-occupancy model from the previous section but will typically be interested in effects of covariates, e.g., to model environmental effects on occupancy. In this section, we show covariate modeling and predictions of occupancy and detection, discuss the difference between the estimate of occupancy probability versus that of the realized occurrence state, and do a bootstrap assessment of uncertainty. We simulate data under the following model:

$z_i \sim Bernoulli(\psi_i)$ with $logit(\psi_i)=\beta_0+\beta_1*vegHt_i$

$y_{ij}|z_i \sim Bernoulli(z_i p_{ij})$

with $logit(p_{ij})=\alpha_0+\alpha_1*wind_{ij}$

Occupancy is affected by a site covariate (vegetation height) and detection is affected by a sampling, or observational covariate (wind speed).

```{r}
# Choose sample sizes and prepare obs. data array y
set.seed(1)                   # So we all get same data set
M <- 100                      # Number of sites
J <- 3                        # Number of presence/absence measurements
y <- matrix(NA, nrow = M, ncol = J) # to contain the obs. data

# Create a covariate called vegHt
vegHt <- sort(runif(M, -1, 1)) # sort for graphical convenience

# Choose parameter values for occupancy model and compute occupancy
beta0 <- 0                    # Logit-scale intercept
beta1 <- 3                    # Logit-scale slope for vegHt
psi <- plogis(beta0 + beta1 * vegHt) # Occupancy probability
par(mfrow = c(1, 3), mar = c(5,5,2,2), cex.axis = 1.5, cex.lab = 1.5)
plot(vegHt, psi, ylim = c(0,1), type = "l", lwd = 3) # Plot psi relationship

# Now visit each site and observe presence/absence perfectly
z <- rbinom(M, 1, psi)        # True presence/absence

# Look at data so far
table(z)

# Plot the true system state
plot(vegHt, z, xlab="Vegetation height", ylab="True presence/absence (z)", frame = F, cex = 1.5)
plot(function(x) plogis(beta0 + beta1*x), -1, 1, add=T, lwd=3, col = "red")
```

There will almost always be false-negative measurement errors. Occurrence z becomes a latent state then, i.e., it will be only partially observable. We simulate this next and imagine that detection probability p is related to the covariate wind via a logit-linear regression with intercept -2 and slope -3 and that we make J=3 presence/absence measurements at each site.

```{r}
# Create a covariate called wind
wind <- array(runif(M * J, -1, 1), dim = c(M, J))

# Choose parameter values for measurement error model and compute detectability
alpha0 <- -2                        # Logit-scale intercept
alpha1 <- -3                        # Logit-scale slope for wind
p <- plogis(alpha0 + alpha1 * wind) # Detection probability
plot(p ~ wind, ylim = c(0,1))     # Look at relationship

# Take J = 3 presence/absence measurements at each site
for(j in 1:J) {
  y[,j] <- rbinom(M, z, p[,j])
}
sum(apply(y, 1, max))               # Number of sites with observed presences

# Plot observed data and true effect of wind on detection probability
plot(wind, y, xlab="Wind", ylab="Observed det./nondetection data (y)", frame = F, cex = 1.5)
plot(function(x) plogis(alpha0 + alpha1*x), -1, 1, add=T, lwd=3, col = "red")

# Look at the data: occupancy, true presence/absence (z), and measurements (y)
cbind(psi=round(psi,2), z=z, y1=y[,1], y2=y[,2], y3=y[,3])
```

We suggest that you look at this table to make sure you really understand the relationships among j (psi), z, and y. Next, we use the site-occupancy model to analyze these data using unmarked and BUGS. We start with unmarked and will also illustrate the fitting of two factors that are unrelated to the data (because the response was not generated with their effects "built in"): time will index the first through the third survey, while hab will contrast three imaginary habitat types.

```{r}
# Create factors
time <- matrix(rep(as.character(1:J), M), ncol = J, byrow = TRUE)
hab <- c(rep("A", 33), rep("B", 33), rep("C", 34))  # Must have M = 100
```

To fit the model in unmarked, we package the data into an unmarked frame first. Note the difference between site covariates (indexed by site only) and sampling or observational covariates (indexed by site and survey). There is really a third possible type of covariate, for time, but in unmarked,thistypehasto be specified as an observational covariate, as we see for factor time, which codes for the first to the third survey.

```{r}
# Load unmarked, format data and summarize
library(unmarked)
umf <- unmarkedFrameOccu( y = y, # Pres/Abs measurements
                          siteCovs = data.frame(vegHt = vegHt, hab = hab), # site-specific covs.
                          obsCovs = list(wind = wind, time = time)) # obs-specific covs.
summary(umf)

# Fit model and extract estimates
# Detection covariates follow first tilde, then occupancy covariates
summary(fm1.occ <- occu(~wind ~vegHt, data=umf))

# Predict occupancy and detection as function of covs (with 95% CIs)
# Add truth from data simulation (below for full code to produce fig. 10-2)
newdat <- data.frame(vegHt=seq(-1, 1, 0.01))
pred.occ <- predict(fm1.occ, type="state", newdata=newdat)
newdat <- data.frame(wind=seq(-1, 1, 0.1))
pred.det <- predict(fm1.occ, type="det", newdata=newdat)

# Predictions for specified values of vegHt, say 0.2 and 2.1
newdat <- data.frame(vegHt=c(0.2, 2.1))
predict(fm1.occ, type="state", newdata=newdat, append = T)

# ... for values of wind of -1 to 1
newdat <- data.frame(wind=seq(-1, 1, .5))
predict(fm1.occ, type="det", newdata=newdat, append = T)
```

We may summarize the analysis by plotting the observed data (the observed occurrence state of every site), the true data-generating values, and the estimated relationship between occupancy and covariate vegHt under the occupancy model (a "logistic regression" that does account for imperfect detection p) and under a simple logistic regression that does not account for p (Figure 10.2 right). We see that ignoring imperfect detection leads to (1) underestimation of the extent of occurrence and (2) to a bias toward zero (attenuation) of the regression coefficient of vegHt.

```{r}
# Fit detection-naive GLM to observed occurrence and plot comparison
summary(fm.glm <- glm(apply(y, 1, max) ~ vegHt, family=binomial))
plot(vegHt, apply(y, 1, max), xlab="Vegetation height", ylab="Observed occurrence ('ever observed ?')", frame = F, cex = 1.5)
plot(function(x) plogis(beta0 + beta1*x), -1, 1, add=T, lwd=3, col = "red")
lines(vegHt, predict(fm.glm, ,"response"), type = "l", lwd = 3)
lines(vegHt, predict(fm1.occ, type="state")[,1], col = "blue", lwd = 3)
legend(-1, 0.9, c("Truth", "'LR' with p", "LR without p"), col=c("red", "blue", "black"), lty = 1, lwd=3, cex = 1.2)
```

The binary random effects zi can be estimated using the function ranef. These random effects have a very tangible meaningdthey are the presence/absence state at each site, and their estimates represent our best guess of whether a particular site is occupied or not.

```{r}
ranef(fm1.occ)
```

These predictions of the random effects z are also called conditional occupancy probability, where conditional means "given the observed data at that site" (MacKenzie et al., 2006, pp. 97e98). When a species has been detected at least once at a site, under the usual assumption of no false-positives the site is occupied with certainty. This is why for sites 91, 92, and 95 in our example the conditional occupancy probability is equal to 1 with zero uncertainty. The case is more interesting for a site where a species was never detected during the J surveys, i.e., {yi} ¼ 0. The probability that site i is occupied then depends on three things: the expected occupancy probability for the site (j), detection probability for the site (p), and the number of surveys J.

This result follows directly from an application of Bayes' rule (Section 2.5.1) and makes sense intuitivelydall else equal, given that the species was not observed at a site, we have higher confidence in its presence despite the negative survey results (1) when it is widespread overall (i.e., when occupancy probability j is high), (2) when it is elusive (i.e., when detection probability p is small), and (3) when the number of times we have looked for it (J) is small.
